### 2023-03-24 Manuscript/Draft1.Rmd
small manuscript changes, many diagnostic changes

The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell $I$ to cell $J$ using the formulation

@@

The following sections describe how we formulated the habitat and temperature components to model real species on the northeast continental shelf.

@@

Species-specific habitat preferences were derived using the *lrren* tool from the R package *envi* [@envi] to create a niche model for each species. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow  We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab^2_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab^2_{J,s}$ for each species.

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

>>>>

The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell $I$ to cell $J$ using the formulation

@@

The package was designed to generate hypothetical temperature gradients and theoretical habitat preferences using Gaussian Random Fields. The following sections describe how we formulated the habitat and temperature components to model real species in the western Atlantic Ocean.

@@

Species-specific habitat preferences were derived from niche model for each species using the *lrren* tool from the R package *envi* [@envi]. The *lrren* tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM [@chen2006unstructured]. The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method **cite arcmap** using point data collected by the United States Geologic Survey (USGS) [@mcmullen2005usgs]. Since the values in $Hab_{J,s}$ are required to be between 0 and 1, we transform the spatial estimates from *lrren* to fall between these bounds. See Figure \@ref(fig:hab-plot1) for a visual representation of this process being applied to Cod. Figure \@ref(fig:hab-plot) depicts habitat preferences $Hab_{J,s}$ for each species.

@@

Each species is assumed to have normally distributed temperature preferences ($N(\mu,\sigma)$). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are $N(8.75,4.25)$, while Haddock and Cod have preferences $N(9,4)$. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM [@chen2006unstructured]. We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \@ref(fig:temp-scenarios) depicts mean trends for the temperature scenarios used in our models. **dont forget to include animated gif in final submission**

### 2023-03-24 Manuscript/Draft1.tex
small manuscript changes, many diagnostic changes

The package was designed to generate theoretical habitat preferences using Gaussian Random Fields that combine with hypothetical temperature gradients to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation

@@

Species-specific habitat preferences were derived using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022) to create a niche model for each species. The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap} using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab^2_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab^2_{J,s}\) for each species.

@@

Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)). We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). We chose these values by combining information in the literature with temperatures recorded in the bottom trawl survey. Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include animated gif in final submission}

>>>>

The package combines species-specific temperature tolerances with habitat preferences to drive the probability of movement from cell \(I\) to cell \(J\) using the formulation

@@

Species-specific habitat preferences were derived from niche model for each species using the \emph{lrren} tool from the R package \emph{envi} (Buller 2022). The \emph{lrren} tool estimates an ecological niche using the relative risk function by relating presence/absence data to two covariate predictors. We used bottom trawl point data in from 2009-2021 as our presence/absence input by using a value of 0 for any tow that failed to catch the given species and weighting a successful catch by the biomass of the given tow. We combined data from both the fall and spring surveys to obscure the influence of temperature so that the niche model would instead infer habitat preferences. Depth and mean sediment size were used as our covariate predictors. Estimated depth for the region was obtained from FVCOM (Chen et al. 2006). The mean sediment size raster was interpolated in ArcMap using the natural neighbor interpolation method \textbf{cite arcmap} using point data collected by the United States Geologic Survey (USGS) (McMullen et al. n.d.). Since the values in \(Hab_{J,s}\) are required to be between 0 and 1, we transform the spatial estimates from \emph{lrren} to fall between these bounds. See Figure \ref{fig:hab-plot1} for a visual representation of this process being applied to Cod. Figure \ref{fig:hab-plot} depicts habitat preferences \(Hab_{J,s}\) for each species.

@@

Each species is assumed to have normally distributed temperature preferences (\(N(\mu,\sigma)\)). Values were chosen by combining information in the literature with temperatures recorded in the bottom trawl survey. We assume Yellowtail Flounder's preferences are \(N(8.75,4.25)\), while Haddock and Cod have preferences \(N(9,4)\). Weekly estimated temperature data for the region for 2012 was obtained from FVCOM (Chen et al. 2006). We chose to repeat estimates for a single year rather than use data for consecutive years to reduce the number of factors impacting model dynamics while still incorporating real data. The 2012 data was chosen because it displayed an average temperature pattern that consistently oscillated between maximum and minimum temperature values, allowing for a smooth repeating yearly temperature pattern for the constant temperature scenario. The 2012 temperature data was also transformed to create an oscillating pattern that increases 5 degrees Celsius on average over the duration of the simulation. We chose a 5 degree increase over a 20 year simulation to allow temperature change to have a meaningful impact on dynamics while remaining within reasonable computational limits in terms of the length of the simulation. Figure \ref{fig:temp-scenarios} depicts mean trends for the temperature scenarios used in our models. \textbf{dont forget to include animated gif in final submission}
